The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-07-28 VegOzone.tex
Edit intro.

\begin{abstract}
  As a result of successful air quality regulations, the past two decades saw a steady reduction in episodic high summer peak concentrations of ozone over Europe and North America. At the same time, tropospheric background ozone levels kept increasing in the northern hemisphere until recently. An increased occurrence of persistent heatwaves, as one possible future scenario, generates favorable conditions for the formation of ambient air ozone. Vegetation, highly specialized to sub-arctic climate, is vulnerable to rapid environmental changes inflicted by global warming and might become more susceptible to ozone in the future. Over large parts of Europe the summer 2018 had been extraordinarily hot and dry and caused large wildfires in northern Sweden in particular. This can be regarded as a test case for a future scenario. In both 2018 and 2019, we have monitored ambient air ozone concentrations at NIBIO Environment Centre Svanhovd in Northern Norway. Over all, ozone concentrations did not differ significantly between the two years. While temperatures and global irradiance diverged significantly from multi annual mean, precipitation varied only to some extend. Coincidentally, we have observed ozone-induced visible injuries on clovers in the ozone garden at Svanhovd in 2018, but not in 2019. We investigate the difference in uptake of ozone using the DO3SE model, with respect to the typical vegetation (beech/birch and conifers) at the location. We assess whether critical levels for these species were breached during the summer seasons. Due to data acquisition problem, ozone concentrations for two weeks in July 2018 were missing from our record and have been reconstructed based on the measurements from the Swedish and Finnish ozone measurement stations in the region.
\end{abstract}

@@

With a radiative forcing of $0.40 \pm 0.20\,\unit{W\,m^{-2}}$, ozone in itself is a potent climate forcer, ranked third behind \chem{CO_2} and \chem{CH_4} \citep[Chapter 8]{IPCC2013}. Its damaging effects on global vegetation might provide an important additional feedback to climate change \citep{Nat:Sitch2007}.

@@

Despite a relatively short tropospheric life-time of roughly $22\,\unit{days}$ ranging from a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}, ozone itself is subject to advection over long distances.
In the annual cycle a distinct maximum typically occurs in spring followed by lower concentrations throughout the summer probably due to a higher uptake on land vegetation than on bare ground or snow and ice \citep{AE:Monks2000}. This spring peak is reached earlier in northern Fennoscandia than in more southern locations but is highly variable throughout the years \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is a complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

@@

The opening of stomata, typically quantified through a conductance to water vapor $g_\mathrm{sto}$, is governed by different environmental factors and is described by a multiplicative ansatz \citep{PTRS:Jarvis1976,ICP:MappingManual2017}:
\begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within a range $0-1$. They are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type (PFT) $k$. Stomata are typically considered closed below a light intensity of $50\,\unit{W\,m^{-2}}$.

To measure the risk of damage inflicted to vegetation through ozone, many different metrics have been defined. Two main types can be differentiated, concentration-based (e.g., $\mathrm{AOT_{40}}$) and flux-based (e.g.,$\mathrm{POD_{0.1}}$) metrics. Due to their simplicity, $\mathrm{AOT_{x}}$ are still widely used, despite the actual integrated flux of ozone entering through the stomata is more relevant for an estimate of the actual toxic dose \citep{ICP:MappingManual2017}. 
The Accumulated Ozone Exposure over a threshold concentration of $\mathrm{x}$ in units of $\unit{ppb}$ is defined as:
\begin{equation}
  \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
  \label{eq:aot}
\end{equation}
with the ambient ozone concentration $\chem{[O_3]}$. The integration time varies, but is usually defined from $8\,\unit{am}-8\,\unit{pm}$.
The Phototoxic Ozone Dose (POD) over a threshold y is defined as:
\begin{equation}
  \mathrm{POD_y} = \int{\mathrm{max(\Phi_\chem{[O_3]}-y)}\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
with the ambient ozone flux through the stomata $\Phi_\chem{[O_3]} = g_\mathrm{sto} \cdot \chem{[O_3]}$ and the flux threshold $\mathrm{y}$. The stomatal conductance is either measured directly or computed in accordance to Eq.~(\ref{eq:stomatal}).\\

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record - favorable conditions for boundary layer ozone formation.

@@

Within the scope of our project to assess probable effects of midnight sun conditions on ozone uptake by plants and resultant enhanced susceptibility to ozone inflicted damage, we conducted ambient air ozone concentration measurements at Svanhovd in the growing seasons of 2018/19 in collaboration with NILU. 
In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden at NIBIO Environment Centre Svanhovd. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.
The respective observed average ozone concentration in 2018/19, however, did not differ significantly by statistical means.

>>>>

\begin{abstract}
  An increased occurrence of persistent heatwaves, as one possible future scenario, generates favorable conditions for the formation of ambient air ozone. Vegetation highly specialized to sub-arctic climate is vulnerable to rapid environmental changes inflicted by global warming and might become more susceptible to ozone in the future. Over large parts of Europe the summer 2018 had been extraordinarily hot and dry and caused large wildfires in northern Sweden in particular. This can be regarded as a test case for such a future scenario. In both 2018 and 2019, we have monitored ambient air ozone concentrations at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd in Northern Norway. Due to data acquisition problems, ozone concentrations for two weeks in July 2018 were missing from our record. We present a reconstruction based on probability density function with respect to the Swedish and Finnish atmospheric monitoring sites in the region. Over all, ozone concentrations did not differ significantly between the two years. While temperatures and global irradiance diverged significantly from multi annual mean, precipitation varied only to some extend. Coincidentally, we have observed ozone-induced visible injuries on clovers in the ozone garden at Svanhovd in 2018, but not in 2019. We investigate the difference in uptake of ozone using the DO3SE model, with respect to the typical vegetation (beech/birch and conifers) at the location. We assess whether critical levels for these species were breached during the summer seasons. {\bf TODO: Summary of results.}
\end{abstract}

@@

With a radiative forcing of $0.40 \pm 0.20\,\unit{W\,m^{-2}}$, ozone is a potent climate forcer, ranked third behind \chem{CO_2} and \chem{CH_4} \citep[Chapter 8]{IPCC2013}. Its damaging effects on global vegetation might provide an important additional feedback to climate change \citep{Nat:Sitch2007}.

@@

Despite a relatively short tropospheric life-time of roughly $22\,\unit{days}$, ranging from a few days in the tropical boundary layer to up to $1\,\unit{year}$ in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}, ozone is subject to advection over long distances.
In the annual cycle, a distinct maximum typically occurs in spring followed by lower concentrations throughout the summer. This is most likely due to stronger dry deposition on land vegetation than on bare ground or snow and ice \citep{AE:Monks2000}. This spring peak is reached earlier in northern Fennoscandia than in more southern locations but is highly variable throughout the years \citep{AB:Klingberg2009, BER:Klingberg2019}. A cause for this is a complex interplay between atmospheric dynamics and chemistry \citep{AE:Laurila1996,BER:Hatakka2003} as well as anthropogenic and biogenic activity \citep{AE:Rummukainen1996,AE:Simpson2002,QJRMS:Galbally2007,NGS:Schnell2009}. \citet{ACP:Andersson2017} showed in a modeling study focused on Fennoscandia that the variability in winter can be attributed mainly to changes in atmospheric background (transport) of ozone, while summertime abundance is mostly affected by emissions of precursors in the rest of Europe. Meteorology plays an important role throughout the year.

@@

The opening of stomata, quantified through their conductance to water vapor $g_\mathrm{sto}$, is governed by different environmental factors and can be described by a multiplicative ansatz \citep{PTRS:Jarvis1976,ICP:MappingManual2017}:
\begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within a range $0-1$. They are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type (PFT) $k$. Stomata are typically considered closed below a light intensity of $50\,\unit{W\,m^{-2}}$.

To measure the risk of damage inflicted to vegetation through ozone, many different metrics have been defined. Two main types can be differentiated, concentration-based (e.g., $\mathrm{AOT_{40}}$) and flux-based (e.g.,$\mathrm{POD_{0.1}}$) metrics. Due to their simplicity, $\mathrm{AOT_{x}}$ are still widely used, despite the actual integrated flux of ozone entering through the stomata is more relevant for an estimation of the actual toxic dose \citep{ICP:MappingManual2017}. 
The Accumulated Ozone Exposure over a threshold concentration of $\mathrm{x}$ in units of $\unit{ppb}$ is defined as:
\begin{equation}
  \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
  \label{eq:aot}
\end{equation}
with the ambient ozone concentration $\chem{[O_3]}$. The integration time varies, but is usually defined from $8\,\unit{am}-8\,\unit{pm}$.
The Phototoxic Ozone Dose (POD) over a threshold y is defined as:
\begin{equation}
  \mathrm{POD_y} = \int{\mathrm{max(\Phi_\chem{[O_3]}-y)}\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
with the ambient ozone flux through the stomata $\Phi_\chem{[O_3]} = g_\mathrm{sto} \cdot \chem{[O_3]}$ and the flux threshold $\mathrm{y}$. The stomatal conductance is either measured directly or computed in accordance to Eq.~(\ref{eq:stomatal}).\\

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for boundary layer ozone formation.

@@

Within the scope of our project to assess probable effects of midnight sun conditions on ozone uptake by plants and resultant enhanced susceptibility to ozone inflicted damage, we conducted ambient air ozone concentration measurements at Svanhovd in the growing seasons of 2018/19 in collaboration with the Norwegian Institute for Air Research (NILU). 
In 2018, we have observed ozone damage on semi-natural vegetation (clover) and crops (tobacco) in the ozone garden at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd. In contrast, no such damage was found on the clovers in 2019, although it was found on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.
The respective observed average ozone concentration in 2018/19, however, did not differ significantly by statistical means.
